Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-05 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2914 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1d6ee5ea7dcf6004aa962e05092f79a10362c5af

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed for probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes, e.g., in air traffic management as considered in this study.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a narro valley within the European Alps.
Topography channels the wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements (lagged by one hour) and their derivatives, such as 
wind direction and wind speed at the airport or spatial and temporal differences.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates used for splitting
  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.
  Note that in the meteorological context wind direction is defined on the 
  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind 
regimes and are consequently associated with a very low estimated concentration 
parameter~$\kappa$, i.e. a high estimated variance. 
In terms of covariates, the lagged wind direction (``persistence'') is mostly 
responsible for distinguishing the broad range of wind regimes listed above while
the pressure gradients and wind speed separate the data into subgroups with high 
vs.\ low precision. 
A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.

@@

The performance of the novel circular trees and forests are assessed in a real case study for shortterm
probabilistic wind direction forecasting at two airports with different site characteristics.
As benchline models, probabilistic climatology and perstistency models, as well as a state-of-the-art
circular GLM are evaluated based on proper scoring rules.
In summary, the circular trees and forests have the highest predictive performance in this setting.
For cases without changes in the wind regime, lagged response values provide already
highly skillful estimates leading to good performance of the persistency model for shortterm wind direction forecasts. While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
Hence, the tree-based models provide reliable forecasts in all tested meterological settings.
For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates. While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods. 
Another possible improvement for obtaining more parsimonious circular trees is
to consider splitting circular covariates into two circle segments by searching
two split points simultaneously rather than sequentially at different depths.
While this might slightly improve the predictive performance of circular trees,
this should not affect the performance  of circular forest, as these
consist of very large trees with many different splits.

To conlcude, the tree structure can capture non-linear changes, shifts, and potential interactions in 
covariates without prespecification of such effects. As supported by the presented case study,
this can be particularly useful for modeling a highly fluctuating response, such as typically 
observed for wind direction, or/and in case of a large number of possible covariates.
Moreover, the case study shows that building ensembles of circular trees can even improve the forecasting performance, as the resulting circular forests allow for modeling smooth effects and stabelize the model.

>>>>

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed for probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes, e.g., in air traffic management as considered in this study.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a narrow valley within the European Alps.
Topography channels the wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements (lagged by one hour) and their derivatives, such as 
wind direction and wind speed at the airport or spatial and temporal differences.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates used for splitting
  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.
  Note that in the meteorological context wind direction is defined on the 
  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind 
regimes and are consequently associated with a very low estimated concentration 
parameter~$\kappa$, i.e. a high estimated variance. 
In terms of covariates, the lagged wind direction (``persistence'') is mostly 
responsible for distinguishing the broad range of wind regimes listed above while
the pressure gradients and wind speed separate the data into subgroups with high 
vs.\ low precision. 
A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.

@@

The performance of the novel circular trees and forests are assessed in a real case study for shortterm
probabilistic wind direction forecasting at two airports with different site characteristics.
As benchline models, probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM are evaluated based on proper scoring rules.
In summary, the circular trees and forests have the highest predictive performance in this setting.
For cases without changes in the wind regime, lagged response values provide already
highly skillful estimates leading to a good performance of the persistency model as observed for shortterm wind direction forecasts in this study. While in these cases the trees and forests also benefit from the highly informative lagged response, they can compensate for a lower information of this covariate by incorporating other quantities and possible interactions of these, contrary to the persistency model (see Figure~\ref{fig:boxplot_crpsskill}).
Hence, the tree-based models provide reliable forecasts in all tested meteorological settings.
For operational use, a possible extension could be the incorporation of numerical weather predictions as (additional) covariates. While this probably only slightly improves the predictive skill for short leadtimes, it possibly extends the potential forecast range of the different methods. 
Another possible improvement for obtaining more parsimonious circular trees is
to consider splitting circular covariates into two circle segments by searching
two split points simultaneously rather than sequentially at different depths.
While this might slightly improve the predictive performance of circular trees,
this should not affect the performance  of circular forests, as these
consist of very large trees with many different splits.

To conclude, in general the tree structure can capture non-linear changes, shifts, and potential interactions in 
covariates without prespecification of such effects. As supported by the presented case study,
this can be particularly useful for modeling a highly fluctuating response, such as typically 
observed for wind direction, or/and in case of a large number of possible covariates.
Moreover, the case study shows that building ensembles of circular trees can even improve the forecasting performance, as the resulting circular forests allow for modeling smooth effects and stabelize the model.
